Kaggle-in-class Data Challenges Can Boost Student Learning

### 2018-05-27 paper-kaggle.Rmd
close to final version

The experiment was conducted during Semester 2 2017. Data was collected during three classes, one at the University of Melbourne (MAST90083), and two at Monash University (ETC2420/5242 and ETC3250).

@@

ETC3250, called Business Analytics, is an undergraduate course focusing on data mining. All students participated in a kaggle competition on the classification challenge. Because this group had no comparison group, it was difficult to assess performance.

@@

The competition performance relative to number of submissions is shown in plots d-f. Each point corresponds to one student, and accuracy or error of the best predictions submitted is used. The regression competition seemed to engage students more than the classification challenge. Students submitted more predictiona, and their models improved with more submissions.

@@

### MAST90083

Response rate was 55%, with 34 of 61 students completed the survey.

```{r likert, fig.caption="Stacked bar charts of responses to survey."}
MAST90083_survey <- read_xlsx("data/Data Competition Feedback MAST90083.xlsx") %>% filter(`Email Address` != "julia.polak@unimelb.edu.au")
MAST90083_likert <- MAST90083_survey %>% 
  rename(Q1=`1. I found the data competition is great fun.`) %>%
  rename(Q2=`2. Taking part in the data competition contributed a lot to my engagement with the subject.`) 
Q1 <- MAST90083_likert %>% count(Q1) %>% filter(!is.na(Q1)) %>%
  mutate(response = fct_relevel(Q1, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q1") %>%
  mutate(pct = n/sum(n))
Q2 <- MAST90083_likert %>% count(Q2) %>% filter(!is.na(Q2)) %>%
  mutate(response = fct_relevel(Q2, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q2")  %>%
  mutate(pct = n/sum(n))
Q <- bind_rows(Q1, Q2)
ggplot(Q, aes(x=question, weight = n, 
              fill=fct_relevel(response, "Strongly Disagree", 
                               "Disagree", "Neutral", "Agree", 
                               "Strongly Agree"))) + 
  xlab("") +
  geom_bar(position="fill") + coord_flip() +
  scale_fill_manual("", values=c("#CA0020","#F4A582","#F7F7F7","#92C5DE", "#0571B0")) +
  theme(legend.position = "bottom")
```

<!--
\begin{table}[ht]
\begin{tabular}{}
\end{table}
-->

```{r eval=FALSE}
# Di's survey
# Check that they agreed to the use of this data, Q2
# Qs
# Q1. I found the data competition is great fun.
# Q2. Taking part in the data competition contributed a lot to my engagement with the subject.
# Q3. Taking part in the data competition improved my confidence in my understanding of the covered material.
# Q4. Taking part in the data competition improved my confidence in my success in the final exam.
# Q5. Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.
# Q6. I feel that the required time investment in the data competition was worthy.
# Q7. What was your favorite topic in this subject during the semester?
# Q8. How heavily have you been involved in the submissions? Rate from 1 to 5 (1-not involved at all, 5- very much involved).
# Q9. Your level of involvement in the submissions, compared to your team members, was
# Q10. How heavily have you been involved in the model development? Rate from 1 to 5 (1-not involved at all, 5- very much involved).
# Q11. Your involvement in the model development, compared to your team members, was
MAST90083_survey <- read_xlsx("data/Data Competition Feedback MAST90083.xlsx") %>% filter(`Email Address` != "julia.polak@unimelb.edu.au")
MAST90083_survey %>% count(`1. I found the data competition is great fun.`)
MAST90083_survey %>% count(`2. Taking part in the data competition contributed a lot to my engagement with the subject.`)
MAST90083_survey %>% count(`3. Taking part in the data competition improved my confidence in my understanding of the covered material.`)
MAST90083_survey %>% count(`4. Taking part in the data competition improved my confidence in my success in the final exam.`)
MAST90083_survey %>% count(`5. Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.`)
MAST90083_survey %>% count(`6. I feel that the required time investment in the data competition was worthy.`)
MAST90083_survey$`7. What was your favorite topic in this subject during the semester?`
MAST90083_survey %>% count(`8. How heavily have you been involved in the submissions? Rate from 1 to 5 (1-not involved at all, 5- very much involved).` )
MAST90083_survey %>% count(`9. Your level of involvement in the submissions, compared to your team members, was`)
MAST90083_survey %>% count(`10. How heavily have you been involved in the model development? Rate from 1 to 5 (1-not involved at all, 5- very much involved).` )
MAST90083_survey %>% count(`11. Your involvement in the model development, compared to your team members, was`)
```

### ETC5242

The response rate was 50%, 17 students out of 34 completed the survey.

```{r eval=FALSE}
# Di's survey
# Check that they agreed to the use of this data, Q2
# Qs
# Q1. I found the data competition is great fun.
# Q2. Taking part in the data competition contributed a lot to my engagement with the subject.
# Q3. Taking part in the data competition improved my confidence in my understanding of the covered material.
# Q4. Taking part in the data competition improved my confidence in my success in the final exam.
# Q5. Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.
# Q6. I feel that the required time investment in the data competition was worthy.
# Q7. What was your favorite topic in this subject during the semester?
# Q8. How heavily have you been involved in the submissions? Rate from 1 to 5 (1-not involved at all, 5- very much involved).
# Q9. Your level of involvement in the submissions, compared to your team members, was
# Q10. How heavily have you been involved in the model development? Rate from 1 to 5 (1-not involved at all, 5- very much involved).
# Q11. Your involvement in the model development, compared to your team members, was
ETC5242_survey <- read_xlsx("data/Data Competition Feedback ETC Di.xlsx") %>% filter(Name != "Dianne Cook")
ETC5242_survey %>% count(`1. I found the data competition is great fun.`)
ETC5242_survey %>% count(`2. Taking part in the data competition contributed a lot to my engagement with the subject.`)
ETC5242_survey %>% count(`3. Taking part in the data competition improved my confidence in my understanding of the covered material.`)
ETC5242_survey %>% count(`4. Taking part in the data competition improved my confidence in my success in the final exam.`)
ETC5242_survey %>% count(`5. Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.`)
ETC5242_survey %>% count(`6. I feel that the required time investment in the data competition was worthy.`)
ETC5242_survey$`7. What was your favorite topic in this subject during the semester?`
ETC5242_survey %>% count(`8. How heavily have you been involved in the submissions? Rate from 1 to 5 (1-not involved at all, 5- very much involved).` )
ETC5242_survey %>% count(`9. Your level of involvement in the submissions, compared to your team members, was`)
ETC5242_survey %>% count(`10. How heavily have you been involved in the model development? Rate from 1 to 5 (1-not involved at all, 5- very much involved).` )
ETC5242_survey %>% count(`11. Your involvement in the model development, compared to your team members, was`)

### ETC3250

```{r eval=FALSE}
# Di's survey
# Check that they agreed to the use of this data, Q2
# Qs
# Q1. I found the data competition is great fun.
# Q2. Taking part in the data competition contributed a lot to my engagement with the subject.
# Q3. Taking part in the data competition improved my confidence in my understanding of the covered material.
# Q4. Taking part in the data competition improved my confidence in my success in the final exam.
# Q5. Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.
# Q6. I feel that the required time investment in the data competition was worthy.
# Q7. What was your favorite topic in this subject during the semester?
# Q8. How heavily have you been involved in the submissions? Rate from 1 to 5 (1-not involved at all, 5- very much involved).
# Q9. Your level of involvement in the submissions, compared to your team members, was
# Q10. How heavily have you been involved in the model development? Rate from 1 to 5 (1-not involved at all, 5- very much involved).
# Q11. Your involvement in the model development, compared to your team members, was
ETC3250_survey <- read_xlsx("data/Data Competition Feedback ETC3250.xlsx") %>% filter(`Email Address` != "dicook@monash.edu") %>% filter(`Email Address` != "souhaib.bentaieb@monash.edu")
ETC3250_survey %>% count(`1. I found the data competition is great fun.`)
ETC3250_survey %>% count(`2. Taking part in the data competition contributed a lot to my engagement with the subject.`)
ETC3250_survey %>% count(`3. Taking part in the data competition improved my confidence in my understanding of the covered material.`)
ETC3250_survey %>% count(`4. Taking part in the data competition improved my confidence in my success in the final exam.`)
ETC3250_survey %>% count(`5. Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.`)
ETC3250_survey %>% count(`6. I feel that the required time investment in the data competition was worthy.`)
ETC3250_survey$`7. What was your favorite topic in this subject during the semester?`
ETC3250_survey %>% count(`8. How heavily have you been involved in the submissions? Rate from 1 to 5 (1-not involved at all, 5- very much involved).` )
ETC3250_survey %>% count(`9. Your level of involvement in the submissions, compared to your team members, was`)
ETC3250_survey %>% count(`10. How heavily have you been involved in the model development? Rate from 1 to 5 (1-not involved at all, 5- very much involved).` )
ETC3250_survey %>% count(`11. Your involvement in the model development, compared to your team members, was`)
```

@@

This paper has discussed results from an experiment to examine the effectiveness of data competitions on student learning, using Kagggle InClass as the vehicle for conducting the competition. The evidence suggests that participating in competitions enhances learning.

@@

This document was produced in R [@R] with the package knitr [@knitr]. Data cleaning was conducted using tidyr [@tidyr], dplyr [@dplyr] and plots were made with ggplot2 [@ggplot2].

# Supplementary material

The following material is provided in addition to the main paper. 
- Code to reproduce the results is in the Rmd document
- De-identified data 
- Additional details of analysis
- Copies of relevant exam questions

>>>>

The experiment was conducted during Semester 2 2017. Data was collected during three classes, one at the University of Melbourne (MAST90083), and one at Monash University (ETC2420/5242).

@@

<!--
ETC3250, called Business Analytics, is an undergraduate course focusing on data mining. All students participated in a kaggle competition on the classification challenge. Because this group had no comparison group, it was difficult to assess performance. 
-->

@@

The competition performance relative to number of submissions is shown in plots (d)-(f). Each point corresponds to one student, and accuracy or error of the best predictions submitted is used. The regression competition seemed to engage students more than the classification challenge. Students submitted more predictiona, and their models improved with more submissions.

@@

Figure \ref{fig:likert} shows the survey responses related to the kaggle competition, for MAST90083 and ETC5242. The response rate for MAST90083 was 55%, with 34 of 61 students completed the survey. The response rate for ETC5242 was 50%, 17 students out of 34 completed the survey.

```{r likert, fig.cap="\\label{fig:likert}Summary of responses to survey of kaggle competition participants. Overwhelmingly the response to the competition was positive in both classes, especially the questions on enjoyment and engagement in the class, and obtaining practical experience."}
MAST90083_survey <- read_xlsx("data/Data Competition Feedback MAST90083.xlsx") %>% filter(`Email Address` != "julia.polak@unimelb.edu.au")
MAST90083_likert <- MAST90083_survey %>% 
  rename(Q1=`1. I found the data competition is great fun.`) %>%
  rename(Q2=`2. Taking part in the data competition contributed a lot to my engagement with the subject.`)  %>%
  rename(Q3=`3. Taking part in the data competition improved my confidence in my understanding of the covered material.`)  %>%
  rename(Q4=`4. Taking part in the data competition improved my confidence in my success in the final exam.`) %>%
  rename(Q5=`5. Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.`) %>%
  rename(Q6=`6. I feel that the required time investment in the data competition was worthy.`)
Q1 <- MAST90083_likert %>% count(Q1) %>% filter(!is.na(Q1)) %>%
  mutate(response = fct_relevel(Q1, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q1") %>%
  mutate(pct = n/sum(n))
Q2 <- MAST90083_likert %>% count(Q2) %>% filter(!is.na(Q2)) %>%
  mutate(response = fct_relevel(Q2, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q2")  %>%
  mutate(pct = n/sum(n))
Q3 <- MAST90083_likert %>% count(Q3) %>% filter(!is.na(Q3)) %>%
  mutate(response = fct_relevel(Q3, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q3")  %>%
  mutate(pct = n/sum(n))
Q4 <- MAST90083_likert %>% count(Q4) %>% filter(!is.na(Q4)) %>%
  mutate(response = fct_relevel(Q4, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q4")  %>%
  mutate(pct = n/sum(n))
Q5 <- MAST90083_likert %>% count(Q5) %>% filter(!is.na(Q5)) %>%
  mutate(response = fct_relevel(Q5, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q5")  %>%
  mutate(pct = n/sum(n))
Q6 <- MAST90083_likert %>% count(Q6) %>% filter(!is.na(Q6)) %>%
  mutate(response = fct_relevel(Q6, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q6")  %>%
  mutate(pct = n/sum(n))
MAST90083_Q <- bind_rows(Q1, Q2, Q1, Q2, Q3, Q4, Q5, Q6)

ETC5242_survey <- read_xlsx("data/Data Competition Feedback ETC Di.xlsx") %>% filter(Name != "Dianne Cook")
ETC5242_likert <- ETC5242_survey %>% 
  rename(Q1=`1. I found the data competition is great fun.`) %>%
  rename(Q2=`2. Taking part in the data competition contributed a lot to my engagement with the subject.`)  %>%
  rename(Q3=`3. Taking part in the data competition improved my confidence in my understanding of the covered material.`)  %>%
  rename(Q4=`4. Taking part in the data competition improved my confidence in my success in the final exam.`) %>%
  rename(Q5=`5. Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.`) %>%
  rename(Q6=`6. I feel that the required time investment in the data competition was worthy.`)
Q1 <- ETC5242_likert %>% count(Q1) %>% filter(!is.na(Q1)) %>%
  mutate(response = fct_relevel(Q1, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q1") %>%
  mutate(pct = n/sum(n))
Q2 <- ETC5242_likert %>% count(Q2) %>% filter(!is.na(Q2)) %>%
  mutate(response = fct_relevel(Q2, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q2")  %>%
  mutate(pct = n/sum(n))
Q3 <- ETC5242_likert %>% count(Q3) %>% filter(!is.na(Q3)) %>%
  mutate(response = fct_relevel(Q3, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q3")  %>%
  mutate(pct = n/sum(n))
Q4 <- ETC5242_likert %>% count(Q4) %>% filter(!is.na(Q4)) %>%
  mutate(response = fct_relevel(Q4, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q4")  %>%
  mutate(pct = n/sum(n))
Q5 <- ETC5242_likert %>% count(Q5) %>% filter(!is.na(Q5)) %>%
  mutate(response = fct_relevel(Q5, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q5")  %>%
  mutate(pct = n/sum(n))
Q6 <- ETC5242_likert %>% count(Q6) %>% filter(!is.na(Q6)) %>%
  mutate(response = fct_relevel(Q6, "Strongly Disagree", "Disagree", "No Strong Feelings", "Agree", "Strongly Agree")) %>% 
  mutate(response = fct_recode(response, "Neutral" = "No Strong Feelings")) %>%
  mutate(question = "Q6")  %>%
  mutate(pct = n/sum(n))
ETC5242_Q <- bind_rows(Q1, Q2, Q1, Q2, Q3, Q4, Q5, Q6)
MAST90083_Q <- MAST90083_Q %>% mutate(class = "MAST90083")
ETC5242_Q <- ETC5242_Q %>% mutate(class = "ETC5242")
Q <- bind_rows(MAST90083_Q, ETC5242_Q)
Q <- Q %>% mutate(class = fct_relevel(class, "MAST90083", "ETC5242"))
ggplot(Q, aes(x=question, weight = n, 
              fill=fct_relevel(response, "Strongly Disagree", 
                               "Disagree", "Neutral", "Agree", 
                               "Strongly Agree"))) + 
  xlab("") + ylab("") +
  facet_wrap(~class, ncol=2) +
  geom_bar(position="fill") + coord_flip() +
  scale_fill_manual("", values=c("#CA0020","#F4A582","#F7F7F7","#92C5DE", "#0571B0")) + 
  theme_bw() +
  theme(legend.position = "bottom") 
```

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|p{15cm}|}\hline
Label & Question \\\hline
Q1 & I found the data competition is great fun.\\\hline
Q2 & Taking part in the data competition contributed a lot to my engagement with the subject.\\\hline
Q3 & Taking part in the data competition improved my confidence in my understanding of the covered material.\\\hline
Q4 & Taking part in the data competition improved my confidence in my success in the final exam.\\\hline
Q5 & Taking part in the data competition improved my confidence in my ability to use the acquired knowledge in practical applications.\\\hline
Q6 & I feel that the required time investment in the data competition was worthy. \\\hline
\end{tabular}
\end{center}
\caption{Questions asked in the survey of competition participants.}
\label{survey_Q}
\end{table}

@@

This paper has described the setup and results of an experiment to examine the effectiveness of data competitions on student learning, using Kaggle InClass as the vehicle for conducting the competition. The experiment was conducted in the classroom settings as part of the normal teaching of the courses, which imposes limitations on the design. However, with both MAST90083 and ETC2420/5242, either a randomized assignment of students to two topic groups, or a control group was possible, which enabled comparing performance.

The primary finding is that participating in competitions results in statistically significant improvement in the learning of the topic, although the effect size is small.  Secondarily, anecdotally the competitions enhanced interest and engagement in the course.

@@

This document was produced in R [@R] with the package knitr [@knitr]. Data cleaning was conducted using tidyr [@tidyr], dplyr [@dplyr] and plots were made with ggplot2 [@ggplot2]. The materials to reproduce the work are available at https://github.com/dicook/paper-quoll.
